---
title: "Exploratory analysis of heart disease data"
author: "Biswajit Chowdhury"
date: "25/08/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load the required library 

```{r,echo=T, message=FALSE}
library(tidyverse)
library(ggpubr)
theme_set(theme_pubr())
theme_set(theme_bw())
```

# 2. load the dataset and overall summary
```{r}
heart_data<-read.csv("heart.csv")
names(heart_data) <- c("age", "sex", "chest_pain", "resting_bp","cholestrol",
                       "fasting_sugar", "resting_ECG", "max_heart_rate", 
                       "exercise_agina", "oldpeak", "slope", "number_major_vessels",
                       "thal", "target")

head(heart_data,3)
dim(heart_data)
str(heart_data)

summary(heart_data)
```

# 3. Preprocessing the data for exploratory analysis
```{r}
# variable sex is coded 0 and 1
# we want to attach value labels 0=F, 1=M

heart_data$target <- factor(heart_data$target,
      levels = c(0,1),
      labels = c("Normal", "Heart Disease"))

heart_data$sex <- factor(heart_data$sex,
      levels = c(0,1),
      labels = c("F", "M"))

heart_data$slope <- factor(heart_data$slope,
      levels = c(1,2,3),
      labels = c("Upsloping", "Flat", "Douwnsloping"))

heart_data$chest_pain <- factor(heart_data$chest_pain,
      levels = c(1,2,3, 0),
      labels = c("Typical angina ", "Atypical angina ", "Non-anginal pain ", "Asymptomatic"))

heart_data$fasting_sugar <- factor(heart_data$fasting_sugar,
      levels = c(0,1),
      labels = c("False", "TRUE"))

heart_data$resting_ECG <- factor(heart_data$resting_ECG,
      levels = c(0,1, 2),
      labels = c("Normal", "Mild", "Severe"))

heart_data$exercise_agina <- factor(heart_data$exercise_agina,
      levels = c(0,1),
      labels = c("No", "Yes"))

heart_data$number_major_vessels <- factor(heart_data$number_major_vessels,
      levels = c(0,1, 2, 3, 4),
      labels = c("None", "One", "Two", "Three", "Four"))

heart_data$thal <- factor(heart_data$thal,
      levels = c(1,2,3),
      labels = c("Normal", "Fixed defect", "Reversable defect"))


```

# 4. Explore the qunatitative vriables

```{r}
# Presence of heart disease among males and females due to age  

ggboxplot(heart_data, x = "target", y = "age", 
          xlab = "Presence of heart disease", ylab = "Age (years)", 
          title = "Heart disease with age", # title of the graph
          color = "sex", # color by gender
          add = c("jitter", "mean_sd"))+ # show the distribution of the values
          theme(plot.title = element_text(hjust = 0.5))
```

Adult females are more prone to get heart disease compared with men.

```{r}
# Presence of heart disease among males and females due to resting blood pressue  

ggboxplot(heart_data, x = "target", y = "resting_bp", 
          xlab = "Presence of heart disease", ylab = "Resting blood pressue (mm Hg)", 
          title = "Resting blood pressue and heart disease",
          color = "sex", add = c("jitter", "mean_sd")) +
          theme(plot.title = element_text(hjust = 0.5))
```

Men and women with heart disease had almost same resting blood pressure.

```{r}
# Changes of chlestreol level between heart disease and normal people  

ggboxplot(heart_data, x = "target", y = "cholestrol", 
          xlab = "Presence of heart disease", ylab = "Cholestoral level (mg/dl)",
          title = "Relationship between heart disease and serum cholestoral",
          color = "sex", add = c("jitter", "mean_sd")) +
          theme(plot.title = element_text(hjust = 0.5))
```

No change in cholestreol level between normal and heart disease.

```{r}
# Chnages of heart rate between normal and disease 

ggboxplot(heart_data, x = "target", y = "max_heart_rate", 
          xlab = "Presence of heart disease", ylab = "Heart rate", 
          title = "Relationship between heart disease and heart rate",
          color = "sex", add = c("jitter", "mean_sd")) +
          theme(plot.title = element_text(hjust = 0.5))
```

Men with heart disease had increased heart rate.

```{r}
# Changes of old peak among the peopple with heart disease and healthy  

ggboxplot(heart_data, x = "target", y = "oldpeak", 
          xlab = "Presence of heart disease", ylab = "ST wave", 
          title = "Relationship between heart disease and ST wave",
          color = "sex", add = c("jitter", "mean_sd")) +
          theme(plot.title = element_text(hjust = 0.5))
```

Presence of reduced ST wave in people with heart disease.

# 5. Explore the Categorical vriables
 
## Calculate the frequency of categorical variables

```{r}
# Counts for gender categories
table(heart_data$sex)


# Cross classification counts for gender by heart failure

table(heart_data$target, heart_data$sex)
```


## Assess the count of people by heart disease, gender, and fasting sugar
```{r}
# Multidimensional tables based on three or more categorical variables. 

table1 <- table(heart_data$target, heart_data$sex, 
                 heart_data$fasting_sugar)

ftable(table1) # print the results more attractively

```

## Assess the count of people by heart failure, gender, chest pain

```{r}
table1 <- table(heart_data$target, heart_data$sex, 
                 heart_data$chest_pain)

ftable(table1)

```

## Assess the count of people by heart disease, gender, and resting ECG

```{r}
table1 <- table(heart_data$target, heart_data$sex, 
                 heart_data$resting_ECG)

ftable(table1)

```

## Assess the count of people by heart disease, gender, exercise agina, and resting ECG

```{r}
table1 <- table(heart_data$target, heart_data$sex, 
                 heart_data$exercise_agina, heart_data$resting_ECG)

ftable(table1)

```


## Visualize the categorical variables

```{r}

# Here we can assess the number of people diagnosed with chest pain
ggplot(heart_data, aes(x = chest_pain, fill=chest_pain)) +
        geom_bar() +  xlab('Level of chest pain') + ylab("Count") 
```

To make it easy to understand, we can make it in desending order. For this we will make a functions that sort the variables crosponding their total counts

```{r}
# re-order levels
reorder_size <- function(x) {
        factor(x, levels = names(sort(table(x), decreasing = TRUE)))
}

ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar() +
  xlab('Level of chest pain') + ylab("Count") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Changes of chest pain with sex and heart disease

```{r}

ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar() +
  xlab('Level of chest pain') +
  facet_grid(target~ sex) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Number of major vessels with gender and heart disease

```{r}

ggplot(heart_data, aes(x = reorder_size(number_major_vessels), fill=number_major_vessels)) +
        geom_bar() +
  xlab('Number of major vessels') +
  facet_grid(target~ sex) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

### The slope of the peak exercise of ST segment with gender and heart disease

```{r}

ggplot(heart_data, aes(x = reorder_size(slope), fill=slope)) +
        geom_bar() +
  xlab('The slope of the peak exercise ST segment') + ylab(" Count")+
  facet_grid(target~ sex) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Changes of thal with sex and heart disease

```{r}

ggplot(heart_data, aes(x = reorder_size(thal), fill=thal)) +
        geom_bar() +
  xlab('Type of thal') + ylab(" Count")+
  facet_grid(target~ sex) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

# 6. Proportion/Percentage
To do this we simply use the frequency tables produced by table() to the prop.table() function


```{r}

# percentages of gender categories
table1<- table(heart_data$sex)
prop.table(table1)

# percentage of cross classication counts for gender by heart disease

table2<- table(heart_data$target, heart_data$sex)
prop.table(table2)

round(prop.table(table2), 3)*100

#percentage of heart failure by sex, and fasting sugar
table1 <- table(heart_data$target, heart_data$sex, 
                 heart_data$fasting_sugar)

ftable(round(prop.table(table1), 3)*100)

```

## Visualize the data 
```{r}

ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar(aes(y = (..count..)/sum(..count..))) +
  xlab('Level of chest pain') + ylab("Count") +
  
  scale_y_continuous(labels = scales::percent, name = "Percentage") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Chest pain with only one variable heart disease
ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar(aes(y = (..count..)/sum(..count..))) +
  xlab('Level of chest pain') + ylab("Count") +
  scale_y_continuous(labels = scales::percent, name = "Percentage") +
  facet_grid(~ sex) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

# chest pain with multiple valriables sex and heart disease

ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar(aes(y = (..count..)/sum(..count..))) +
  xlab('Level of chest pain') +
  scale_y_continuous(labels = scales::percent, name = "Percentage") +
  facet_grid(target~ sex) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```











